---
title: "Introduction to pwrSim"
author: "pwrSim Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pwrSim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `pwrSim` package provides tools for estimating statistical power through simulation for mixed effects models with interactions. This vignette demonstrates the main functions and typical workflows.

## Installation

```{r eval=FALSE}
# Install from GitHub
# devtools::install_github("hdbt/pwrSim")
library(pwrSim)
```

```{r}
library(pwrSim)
```

## Basic Power Analysis

The main function `simulate_power()` estimates power for a given experimental design. You need to specify:

- Number of subjects (`n_subj`)
- Number of observations per subject (`n_obs_per_subj`)
- Fixed effects (including interactions)
- Random effects (standard deviations)
- Residual standard deviation

### Simple Example with Interaction

Let's estimate power for a study with:
- 30 subjects
- 20 observations per subject
- Two main effects and their interaction
- Random intercepts and slopes

```{r example1, eval=FALSE}
# Define effect sizes
fixed_effects <- c(
  intercept = 5.0,      # Overall mean
  effect1 = 0.5,        # Main effect 1
  effect2 = 0.3,        # Main effect 2
  interaction = 0.4     # Interaction effect
)

# Define random effects (as standard deviations)
random_effects <- c(
  subj_intercept = 1.0,  # Random intercept SD
  subj_slope = 0.5       # Random slope SD
)

# Run power analysis
result <- simulate_power(
  n_subj = 30,
  n_obs_per_subj = 20,
  fixed_effects = fixed_effects,
  random_effects = random_effects,
  residual_sd = 1.5,
  n_sim = 1000,
  seed = 123
)

# View results
print(result)
summary(result)
```

## Interpreting Results

The output shows:
- **Power estimates**: Proportion of simulations where each effect was significant
- **Convergence rate**: Proportion of models that converged successfully
- **Study design parameters**: Sample size and effect specifications

Power of 0.80 (80%) is typically considered adequate for most research.

## Sample Size Determination

Use `calculate_sample_size()` to find the number of subjects needed for a target power level:

```{r sample_size, eval=FALSE}
# Find sample size for 80% power on the interaction effect
ss_result <- calculate_sample_size(
  target_power = 0.80,
  effect_name = "interaction",
  n_obs_per_subj = 20,
  fixed_effects = fixed_effects,
  random_effects = random_effects,
  residual_sd = 1.5,
  n_subj_range = c(20, 60),
  n_sim = 500,
  seed = 456
)

print(ss_result)

# Visualize power curve
plot(ss_result)
```

## Parallel Processing

For large simulations, enable parallel processing:

```{r parallel, eval=FALSE}
result_parallel <- simulate_power(
  n_subj = 30,
  n_obs_per_subj = 20,
  fixed_effects = fixed_effects,
  random_effects = random_effects,
  residual_sd = 1.5,
  n_sim = 5000,
  parallel = TRUE,
  n_cores = 4,
  seed = 789
)
```

## Understanding Effect Sizes

Choosing appropriate effect sizes is crucial:

- **Small effects**: Cohen's d ≈ 0.2
- **Medium effects**: Cohen's d ≈ 0.5
- **Large effects**: Cohen's d ≈ 0.8

For mixed models, effect sizes depend on:
- Between-subject variability (random intercept SD)
- Within-subject variability (residual SD)
- Random slope variability

## Tips for Power Analysis

1. **Use realistic effect sizes**: Base them on pilot data or literature
2. **Consider multiple scenarios**: Test optimistic and conservative estimates
3. **Account for attrition**: Plan for participant dropout
4. **Check convergence**: Low convergence rates suggest model misspecification
5. **Use sufficient simulations**: At least 1000 for stable estimates

## Advanced: Custom Formulas

You can specify custom model formulas:

```{r custom_formula, eval=FALSE}
result_custom <- simulate_power(
  n_subj = 25,
  n_obs_per_subj = 15,
  fixed_effects = fixed_effects,
  random_effects = random_effects,
  formula = "y ~ x1 * x2 + (1 + x1 | subj_id)",
  n_sim = 500
)
```

## Detailed Results

Access simulation details for further analysis:

```{r details, eval=FALSE}
# Get detailed results from each simulation
head(result$detailed_results)

# Calculate confidence intervals for power estimates
power_ci <- t.test(result$detailed_results$interaction_sig)$conf.int
cat("95% CI for interaction power:", power_ci, "\n")
```

## Conclusion

The `pwrSim` package provides flexible tools for power analysis in mixed effects models. Key functions:

- `simulate_power()`: Estimate power for a given design
- `calculate_sample_size()`: Find required sample size

For more information, see the function documentation with `?simulate_power` and `?calculate_sample_size`.
